UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-04-11 ms/main.tex
move some stuff to matmethods, small fixes

\begin{abstract}
{\large Abstract.}
%The abstract is typically 100-150 words, unreferenced. 
Improvements in sequencing have generated massive and diverse genomic datasets of populations. Dimension reduction is necessary to condense and analyze the data. Uniform manifold approximation and projection (UMAP) is a non-linear dimension reduction tool recently made available. Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data. We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry population-scale datasets.
\end{abstract}

@@

Uniform Manifold Approximation and Projection (UMAP) is a dimension reduction technique designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals. The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

@@

\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

@@

\subsubsection*{Regional patterns in the Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} The UKBB provides genotype data on 488,377 individuals along with self-identified ethnic background in a hierarchical tree-structured dictionary. Participants provided ethnic background on two occasions. We used the initial ethnicity after finding minimal differences between the two. The dataset is majority White (88.3\% British, 2.6\% Irish, 3.4\% other), with large populations identifying as Black (1.6\% either African, Caribbean, or other), Asian (1.9\% either Indian, Pakistani, Bangladeshi, or other), Chinese (0.3\%), an other ethnic group (0.8\%), mixed ethnicity (0.6\%), or an unavailable response (0.5\%).

UMAP on the top 10 principal components reveals both continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): The patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of large and multi-ethnic datasets.

@@

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group"; some of these were identified using country of birth in figure~\ref{fig:supp_ukbb_cob} and appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.

@@

Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger populations than PCA. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

>>>>

\begin{abstract}
{\large Abstract.}
%The abstract is typically 100-150 words, unreferenced. 
Improvements in sequencing have generated massive and diverse genomic datasets of populations. Dimension reduction is necessary to condense and analyze the data. Uniform manifold approximation and projection (UMAP) is a non-linear dimension reduction tool recently made available. Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data. We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
\end{abstract}

@@

Uniform Manifold Approximation and Projection (UMAP) is a dimension reduction technique designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a patchwork of low-dimensional representations of neighbourhoods that groups genetically similar individuals together on a local scale while better preserving long-range topological connections to more distantly related individuals. The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

@@

\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 Americans from a variety of backgrounds. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

@@

\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region. One cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding. The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB is a trove of data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of large and multi-ethnic datasets.

@@

The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations. The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities. Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group". Using data on countries of birth, we identified many groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.

@@

Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger sample sizes than PCA. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.

The UKBB provides genotype data on 488,377 individuals along with self-identified ethnic background in a hierarchical tree-structured dictionary. Participants provided ethnic background on two occasions. We used the initial ethnicity after finding minimal differences between the two. The dataset is majority White (88.3\% British, 2.6\% Irish, 3.4\% other), with large populations identifying as Black (1.6\% either African, Caribbean, or other), Asian (1.9\% either Indian, Pakistani, Bangladeshi, or other), Chinese (0.3\%), an other ethnic group (0.8\%), mixed ethnicity (0.6\%), or an unavailable response (0.5\%).
